Reducing the Memory Cost of Training Convolutional Neural Networks by CPU Offloading

### 2019-07-09 article.tex
Spell check

In recent years, Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
implementations to get optimal performance from the hardware. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
GPU memory is a major bottleneck of the CNN training procedure,
 limiting  the size of both inputs and model architectures.
%The backpropagation algorithm requires to accumulate the activation 
%values of hidden layers in live GPU memory during the forward pass
%as these activations are needed for the computation of the weights 
%gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
leveraging an under-utilized resource of modern systems: 
the device to host bandwidth.
%offloading the input activation of hidden layers to the CPU:
Our method, termed CPU offloading, works by transfering hidden activations to the CPU upon
computation, in order to free GPU memory for upstream layer computations during the forward pass. 
These activations are then transfered back to the GPU as needed by the gradient computations of the backward pass.
The key challenge to our method is to efficiently overlap data transfers and computations
in order to minimize  wall time overhead induced by the additional data transfers.
On a typical work station with a Nvidia Titan X GPU and XXX PCIe lanes, 
we show that our method compares favorably to 
gradient checkpointing as we are able reduce the memory consumption of traning a VGG19 model by 35\% with a minimal additional wall time overhead of 21\%. 
Further experiments detail the impact of the different optimization tricks we propose.
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that they can easily be combined for further optimizations.

@@

% Our approach
Instead, our approach reduces the GPU memory consumption without introducing any additional
computation by leveraging an under-utilized resource: host-device communication.
We propose to temporarily offload GPU memory buffers to the CPU during the forward pass 
of the computation, and transfering these memory buffers back into GPU memory as needed by the gradient computations during the backward of the backpropagation algorithm.

@@

% Summary
The remainer of this paper is organized as follows: In Section 2, we briefly review the literature for related work.
Section 3 introduces the preliminary notions necessary to understand the root of the GPU memory bottleneck.
Section 4 presents our approach and details the different tricks needed for efficient implementation. 
Finally, Section 5 presents the results of our evaluation.

@@

while Rectified Linear Unit layers need the input activations values 
to compute the gradients of the loss with respect to its imputs:

@@

% Illustrate with figure
Figure 1 illustrates our approach. 
During the forward pass (top), activations are computed forward through the network layers.
Instead of keeping these activations idle in GPU memory, activation values are transfered to 
the CPU memory immediately after their computation.
In the backward pass (bottom), gradients are backpropagated backward 
through the network layers following equations (3).
Our implementation synchronizes the transfer of the layers input activations back to GPU
right before they are needed for their layer's gradient computation.
Hence the key challenge in our implementation consists in synchronizing the data
transfers with the computations so that only the minimal amount of activation values 
is loaded in GPU memory at any given time, while the least amount of time is spent 
waiting for the data transfer.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure1.jpg}
\caption{Illustration of CPU Oflloading. 
During the forward pass (top), 
hidden activation buffers are transfered to CPU memory upon computation.
During the backward pass (bottom), hidden activation buffers 
are transfered back to GPU memory just in time to compute the 
weight gradients.}
\end{figure}

@@

The key challenge in this parallelization scheme consists in efficiently
managaing memory allocation and data transfer so as to
minimize the time $\mathcal{T}_{idle}$ spent awaiting for data transfer.
In this paper, we adopt a simple parallelization strategy:
During the forward pass, activations $z_i$ are transfered to CPU 
upon computation by layer $i$ as $z_i = f_i(z_{i-1})$,
The GPU memory buffer of $z_i$ must then await for both the data transfer to CPU and
the next layer computation $z_{i+1} = f_{i+1}(z_{i})$ to complete before deallocation in order to
avoid unfortunate overwriting by concurrent operations.

\begin{figure*}[h]
\begin{algorithm}[H]
	\KwData{Layer $f_i$, input activation $z_{i-1}$ 
		      \\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		      \\ \hspace{1.1cm} CPU thread $T_{data}$
		      \\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{comp}^i$
	          \\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$}
\vspace{1mm}
	\KwResult{$z_i$}
\vspace{1mm}
    \textit{Allocate}($z_i$)\;
	$S_{comp} \Leftarrow z_i \leftarrow f_i(z_{i-1})$\;
	$S_{comp} \Leftarrow E_{comp}^i$\;
\vspace{1mm}
	\textbf{In Thread}  $T_{data}$:\\
		\hspace{1.1cm} 	$S_{data} \Leftarrow P_{i-1} \leftarrow z_{i-1}$\;
		\hspace{1.1cm}  $S_{data} \Leftarrow E_{data}^i$\;
		\hspace{1.1cm}  \textit{Wait}($E_{data}^i$, $E_{comp}^i$)\;
		\hspace{1.1cm}  \textit{Free}($z_{i-1}$)\;
	\caption{Forward procedure through layer $i$ with parallel CPU offloading. }
\end{algorithm}
\small{Algorithm 1: Double arrows indicate the asynchronous execution of a CUDA directive within a stream.
Data transfers are executed within dedicated CUDA stream and CPU thread to
synchronize the memory deallocation without blocking the execution of upward layers}
\end{figure*}

One important exception to this rule concerns skip connections, as illustrated in Figure 3.
Skip connections induce a delay in the GPU buffer dealocation as the input of residual blocks must
be kept in memory until the end of the residual block computation to be added to the output.
In ResNet architectures, this delay is short enough to have little impact on the memory/wall time trade off.
However, this means that our method is poorly suited to densely connected 
networks such as DenseNet \cite{huang2017densely} or UNet \cite{ronneberger2015u}, which features very long skip connections.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{Figure3.png}
\caption{Illustration of the delay in memory deallocation induced by residual connections.
(left) without residual connection, the input $z_{i-1}$ of layer $f_i$ can be freed from GPU memory 
as $f_i$ computation terminates. 
(right) Residual connections induce a delay in memory deallocation
as $z_{i-1}$ must be kept in GPU memory to be added to the ouput of the residual block. 
}
\end{figure}

@@

% Pinned Memory
CPU memory can be accessed through either pageable memory 
or pinned memory address spaces.
Pinned memory allows for Direct Memory Access (DMA), 
which can significantly speedup data transfer between CPU and GPU devices through PCIe lanes.
However, allocating pinned memory buffers is a time consuming operation, as illustrated in Table 1.
Hence, naively allocating pinned memory buffers within each forward pass would
slow down training due to the overhead of pinned memory allocation.
Instead, we propose to allocate the pinned memory buffers beforehand, 
during the initialization of the CNN.
Our implementation scans the network architecture upon instantiation
and allocates dedicated pinned memory buffers for each 
target layer activations (convolutions and max-pooling). 
During the forward pass, we transfer the GPU data directly into the pre-allocated pinned memory buffer,
and, during the backward pass, we transfer the activations back to GPU from these pinned buffers.
Table 1 quantifies the gains in transfer speed brought by the use of pinned memory buffers, and in Section5, we evaluate the impact of using pinned memory buffers on the overall efficiency of our method.

@@

% Compression
Another way to reduce the time of data transfer between CPU and GPU is to minize 
$\mathcal{V}$ by compressing the hidden layer activations 
before transferring them between devices.
Compression can be achieved by diverse means including sparsification, 
quantization or low rank factorization techniques.
Low-rank factorization would induce non-negligible additional computational costs, 
which would slow down the computation.
Either sparsifying or quantizing the layers activations are interesting solutions.
However, compression would incur minor losses of precision in the activation values,
which might affect the training dynamics.
Although previous works suggest that such minor losses in precision does not negatively impact training,
investigation of compression techniques would require a detailed analysis 
of the impact of compression on accuracy, which is out of the scope of this paper.
Hence we leave this question open for future work as an interesting direction for further optimization.

@@

% Describe system setup.
We conduct our experiments on a workstation with an NVIDIA Titan X GPU connected 
to the host through XXX PCIe lanes, which represents a peak bandwidth of XXXB/s.
We believe this set-up to be representative of a typical workstation for individual practioners, 
if not slightly unfavorable, as our device to host bandwidth is rather limited.
Unless specified otherwise, we condut our experiments on a batch of 32 RGB images 
with a resolution of 128 by 128 pixels.
For each experiment, we are interested in the total memory usage and Wall time of
the computation for a full training iteration including the forward and backward pass through the network.
It is important to note that our method does not modify the actual computations performed
by the network so that the training dynamics and accuracy are unaffected by CPU Offloading.
Hence, we focus on reporting the memory vs. wall time trade-off,
 as the accuracy reached by the model is strictly unaffected.

@@

Our implementation allows to control for the memory foot-print vs. computation wall time trade-off by 
either using more agressive parallelization schemes,
or by offloading different subsets of hidden layers input activations.
In this section, we use the parallelization scheme described in Section 4.2
and explore the memory vs. time trade-off by off-loading different subsets of the model's layers.

@@

Both ResNet architectures and MobileNet \cite{howard2017mobilenets} feature batch normalization layers, 
for which we used the invertible operation of \cite{rota2018place}. 
Invertible batch normalization layers further reduce the memory consumption 
with little time overhead, which explains the slightly better trade-off of ResNet architectures.
On the other hand, the MobileNet architecture uses depth-wise convolutions instead of the 
reguar convolution layers used in both VGG and ResNet architectures.
Depth-wise convolutions have lower arithmetic intensity \cite{wu2018shift} than convolution layers,
i.e.; for input activations of similar size, they perform less, hence faster, operations. 
As computations are performed faster while the data transfer time remains constant,
CPU offloading of depth-wise convolutions incur a more important overhead.
This result highlights the fact that CPU offloading works best for computationally
intensive layers.

@@

The memory/time trade-off achievable by CPU offloading is a function of the ratio 
between computation and data tranfer speed. 
This ratio varies with the specific hardware
being run on, the algorithm used for the convolutions, 
as well the efficiency of their implementation.

On the software side, the experiments presented in this paper were run 
with state-of-the art cudnn implementations of the winograd convolution 
algorithm through the PyTorch framework \cite{paszke2017automatic}.
On the hardware side, we ran experiments on a single 
Nvidia Titan X GPU connected to the host via a 6GB/s bandwitdth connection, 
which we believe to be representative of a typical deep learning student/researcher's workstation.

The exact trade-offs achieved will differ depending on the specific hardware and software used in 
different systems, as well as the parallelization strategy used for CPU offloading. 
However, the fact that data transfers between CPU and GPU can be efficiently parallelized with the GPU computations
means that memory gains can be achieved with minimal time overhead 
on any sytem given an appropriate parallelization strategy.

>>>>

In recent years, Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
implementations to get optimal performance from the hardware. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
GPU memory is a major bottleneck of the CNN training procedure,
 limiting  the size of both inputs and model architectures.
%The backpropagation algorithm requires to accumulate the activation 
%values of hidden layers in live GPU memory during the forward pass
%as these activations are needed for the computation of the weights 
%gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
leveraging an under-utilized resource of modern systems: 
the device to host bandwidth.
%offloading the input activation of hidden layers to the CPU:
Our method, termed CPU offloading, works by transferring hidden activations to the CPU upon
computation, in order to free GPU memory for upstream layer computations during the forward pass. 
These activations are then transferred back to the GPU as needed by the gradient computations of the backward pass.
The key challenge to our method is to efficiently overlap data transfers and computations
in order to minimize  wall time overhead induced by the additional data transfers.
On a typical work station with a Nvidia Titan X GPU and XXX PCIe lanes, 
we show that our method compares favorably to 
gradient checkpointing as we are able reduce the memory consumption of training a VGG19 model by 35\% with a minimal additional wall time overhead of 21\%. 
Further experiments detail the impact of the different optimization tricks we propose.
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that they can easily be combined for further optimizations.

@@

% Our approach
Instead, our approach reduces the GPU memory consumption without introducing any additional
computation by leveraging an under-utilized resource: host-device communication.
We propose to temporarily offload GPU memory buffers to the CPU during the forward pass 
of the computation, and transferring these memory buffers back into GPU memory as needed by the gradient computations during the backward of the backpropagation algorithm.

@@

% Summary
The remainder of this paper is organized as follows: In Section 2, we briefly review the literature for related work.
Section 3 introduces the preliminary notions necessary to understand the root of the GPU memory bottleneck.
Section 4 presents our approach and details the different tricks needed for efficient implementation. 
Finally, Section 5 presents the results of our evaluation.

@@

while Rectified Linear Unit layers need the input activations values 
to compute the gradients of the loss with respect to its inputs:

@@

% Illustrate with figure
Figure 1 illustrates our approach. 
During the forward pass (top), activations are computed forward through the network layers.
Instead of keeping these activations idle in GPU memory, activation values are transferred to 
the CPU memory immediately after their computation.
In the backward pass (bottom), gradients are backpropagated backward 
through the network layers following equations (3).
Our implementation synchronizes the transfer of the layers input activations back to GPU
right before they are needed for their layer's gradient computation.
Hence the key challenge in our implementation consists in synchronizing the data
transfers with the computations so that only the minimal amount of activation values 
is loaded in GPU memory at any given time, while the least amount of time is spent 
waiting for the data transfer.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure1.jpg}
\caption{Illustration of CPU Oflloading. 
During the forward pass (top), 
hidden activation buffers are transferred to CPU memory upon computation.
During the backward pass (bottom), hidden activation buffers 
are transferred back to GPU memory just in time to compute the 
weight gradients.}
\end{figure}

@@

The key challenge in this parallelization scheme consists in efficiently
managing memory allocation and data transfer so as to
minimize the time $\mathcal{T}_{idle}$ spent awaiting for data transfer.
In this paper, we adopt a simple parallelization strategy:
During the forward pass, activations $z_i$ are transferred to CPU 
upon computation by layer $i$ as $z_i = f_i(z_{i-1})$,
The GPU memory buffer of $z_i$ must then await for both the data transfer to CPU and
the next layer computation $z_{i+1} = f_{i+1}(z_{i})$ to complete before de-allocation in order to
avoid unfortunate overwriting by concurrent operations.

\begin{figure*}[h]
\begin{algorithm}[H]
	\KwData{Layer $f_i$, input activation $z_{i-1}$ 
		      \\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		      \\ \hspace{1.1cm} CPU thread $T_{data}$
		      \\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{comp}^i$
	          \\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$}
\vspace{1mm}
	\KwResult{$z_i$}
\vspace{1mm}
    \textit{Allocate}($z_i$)\;
	$S_{comp} \Leftarrow z_i \leftarrow f_i(z_{i-1})$\;
	$S_{comp} \Leftarrow E_{comp}^i$\;
\vspace{1mm}
	\textbf{In Thread}  $T_{data}$:\\
		\hspace{1.1cm} 	$S_{data} \Leftarrow P_{i-1} \leftarrow z_{i-1}$\;
		\hspace{1.1cm}  $S_{data} \Leftarrow E_{data}^i$\;
		\hspace{1.1cm}  \textit{Wait}($E_{data}^i$, $E_{comp}^i$)\;
		\hspace{1.1cm}  \textit{Free}($z_{i-1}$)\;
	\caption{Forward procedure through layer $i$ with parallel CPU offloading. }
\end{algorithm}
\small{Algorithm 1: Double arrows indicate the asynchronous execution of a CUDA directive within a stream.
Data transfers are executed within dedicated CUDA stream and CPU thread to
synchronize the memory de-allocation without blocking the execution of upward layers}
\end{figure*}

One important exception to this rule concerns skip connections, as illustrated in Figure 3.
Skip connections induce a delay in the GPU buffer de-allocation as the input of residual blocks must
be kept in memory until the end of the residual block computation to be added to the output.
In ResNet architectures, this delay is short enough to have little impact on the memory/wall time trade off.
However, this means that our method is poorly suited to densely connected 
networks such as DenseNet \cite{huang2017densely} or UNet \cite{ronneberger2015u}, which features very long skip connections.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{Figure3.png}
\caption{Illustration of the delay in memory de-allocation induced by residual connections.
(left) without residual connection, the input $z_{i-1}$ of layer $f_i$ can be freed from GPU memory 
as $f_i$ computation terminates. 
(right) Residual connections induce a delay in memory de-allocation
as $z_{i-1}$ must be kept in GPU memory to be added to the ouput of the residual block. 
}
\end{figure}

@@

% Pinned Memory
CPU memory can be accessed through either pageable memory 
or pinned memory address spaces.
Pinned memory allows for Direct Memory Access (DMA), 
which can significantly speedup data transfer between CPU and GPU devices through PCIe lanes.
However, allocating pinned memory buffers is a time consuming operation, as illustrated in Table 1.
Hence, naively allocating pinned memory buffers within each forward pass would
slow down training due to the overhead of pinned memory allocation.
Instead, we propose to allocate the pinned memory buffers beforehand, 
during the initialization of the CNN.
Our implementation scans the network architecture upon instantiation
and allocates dedicated pinned memory buffers for each 
target layer activations (convolutions and max-pooling). 
During the forward pass, we transfer the GPU data directly into the pre-allocated pinned memory buffer,
and, during the backward pass, we transfer the activations back to GPU from these pinned buffers.
Table 1 quantifies the gains in transfer speed brought by the use of pinned memory buffers and, in Section 5, we evaluate the impact of using pinned memory buffers on the overall efficiency of our method.

@@

% Compression
Another way to reduce the time of data transfer between CPU and GPU is to minimize 
$\mathcal{V}$ by compressing the hidden layer activations 
before transferring them between devices.
Compression can be achieved by diverse means including sparsification, 
quantization or low rank factorization techniques.
Low-rank factorization would induce non-negligible additional computational costs, 
which would slow down the computation.
Either sparsifying or quantizing the layers activations are interesting solutions.
However, compression would incur minor losses of precision in the activation values,
which might affect the training dynamics.
Although previous works suggest that such minor losses in precision does not negatively impact training,
investigation of compression techniques would require a detailed analysis 
of the impact of compression on accuracy, which is out of the scope of this paper.
Hence we leave this question open for future work as an interesting direction for further optimization.

@@

% Describe system setup.
We conduct our experiments on a workstation with an NVIDIA Titan X GPU connected 
to the host through XXX PCIe lanes, which represents a peak bandwidth of XXXB/s.
We believe this set-up to be representative of a typical workstation for individual practitioners, 
if not slightly unfavorable, as our device to host bandwidth is rather limited.
Unless specified otherwise, we conduct our experiments on a batch of 32 RGB images 
with a resolution of 128 by 128 pixels.
For each experiment, we are interested in the total memory usage and Wall time of
the computation for a full training iteration including the forward and backward pass through the network.
It is important to note that our method does not modify the actual computations performed
by the network so that the training dynamics and accuracy are unaffected by CPU Offloading.
Hence, we focus on reporting the memory vs. wall time trade-off,
 as the accuracy reached by the model is strictly unaffected.

@@

Our implementation allows to control for the memory foot-print vs. computation wall time trade-off by 
either using more aggressive parallelization schemes,
or by offloading different subsets of hidden layers input activations.
In this section, we use the parallelization scheme described in Section 4.2
and explore the memory vs. time trade-off by off-loading different subsets of the model's layers.

@@

Both ResNet architectures and MobileNet \cite{howard2017mobilenets} feature batch normalization layers, 
for which we used the invertible operation of \cite{rota2018place}. 
Invertible batch normalization layers further reduce the memory consumption 
with little time overhead, which explains the slightly better trade-off of ResNet architectures.
On the other hand, the MobileNet architecture uses depth-wise convolutions instead of the 
regular convolution layers used in both VGG and ResNet architectures.
Depth-wise convolutions have lower arithmetic intensity \cite{wu2018shift} than convolution layers,
i.e.; for input activations of similar size, they perform less, hence faster, operations. 
As computations are performed faster while the data transfer time remains constant,
CPU offloading of depth-wise convolutions incur a more important overhead.
This result highlights the fact that CPU offloading works best for computationally
intensive layers.

@@

The memory/time trade-off achievable by CPU offloading is a function of the ratio 
between computation and data transfer speed. 
This ratio varies with the specific hardware
being run on, the algorithm used for the convolutions, 
as well the efficiency of their implementation.

On the software side, the experiments presented in this paper were run 
with state-of-the art cudnn implementations of the winograd convolution 
algorithm through the PyTorch framework \cite{paszke2017automatic}.
On the hardware side, we ran experiments on a single 
Nvidia Titan X GPU connected to the host via a 6GB/s bandwidth connection, 
which we believe to be representative of a typical deep learning student/researcher's workstation.

The exact trade-offs achieved will differ depending on the specific hardware and software used in 
different systems, as well as the parallelization strategy used for CPU offloading. 
However, the fact that data transfers between CPU and GPU can be efficiently parallelized with the GPU computations
means that memory gains can be achieved with minimal time overhead 
on any system given an appropriate parallelization strategy.
